---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Insert your name here"
date: "Insert date here"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

```{r}
staff <- read_csv("data/instructional-staff.csv")
```
```{r}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
```

```{r}
staff_long %>%
  mutate(
    year = as.numeric(year),
    value = as.numeric(value),
    faculty_type = as.factor(faculty_type)
  ) %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line(size = 1) +
  labs(
    title = "Number of Employted Faculty Types (1975-2011)",
    x = "Year",
    y = "Number of Faculty",
    color = "Faculty Type"
  ) +
  theme_minimal()
```

I feel as if as it is, the graph is already showing that the number of part time faculty is going up compared to other faculty types; in fact, it is the only one showing an upward trend.

### Exercise 1

Remove this text, and add your answer for Exercise 1 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 2

Remove this text, and add your answer for Exercise 1 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 3

...

Add exercise headings as needed.
